---
title: "Ag Carpentry - Trial Data"
author: "Brittani"
date: "10/19/2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
opts_knit$set(root.dir = "~/Box/Data Carpentry for Agronomy/DIFM Code and Information/hord f98")
boundary <- read_sf("~/Box/Data Carpentry for Agronomy/DIFM Code and Information/hord f98/boundary.gpkg")
```
####Motivating Questions:
- How can my results help me choose an optimal application rate?

####Objectives with Agricultural Data Types
- Visualize relationship between yield and input variables
- Understand basic principles of linear regression
- Create a linear model relating yield to nitrogen
- Estimate an optimal nitrogen application rate
- Use ggplot to graph your estimated profit and yield response to nitrogen
- Understand the use of a confidence interval for comparing the treatments
- Consider statistical significance of profit differences between nitrogen rates

####Keypoints:
- sf is prefereable for data analysis; it is easier to access the dataframe
- Projecting your data in utm is necessary for many of the geometric operations you perform (e.g. making trial grids and splitting plots into subplot data)
- Compare different data formats, such as gpkg, shp(cpg,dbf,prj,sbn,sbx),geojson,tif

Below are the packages that we will use in this episode.

```{r}
library(sf)
library(fasterize)
library(raster)
library(rjson)
library(httr)
library(rgdal)
library(rgeos)
library(maptools)
library(knitr)
require(tmap)
require(ggplot2)
require(gridExtra)
library(daymetr)
library(readr)
library(measurements)
library(FedData)
```
