---
title: "Introduction to R and RStudio"
questions:
- "How can I work with a spreadsheet in a more advanced and reproducible manner than using MS Excel?"
- "How can I keep my data and analysis organized?"
- "How can I visualize my data with graphs?"
- "Why would basic programming skills be useful to me for these tasks?"
objectives:
- "Describe the purpose of each pane in the RStudio IDE"
- "Create a project to manage scripts and data"
- "Assign a value to a variable"
- "Perform arithmetic"
- "Use a function"
- "View the help page for a function"
- "Create a new function"
- "Import data from spreadsheet software"
- "Install packages"
- "Make a scatter plot using ggplot2"
source: Rmd
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting to know R and RStudio

Blah blah

If you're already proficient with Excel, you will probably already know how to
accomplish a lot of what we are doing in this episode using Excel instead of
R. One reason why we are teaching you this material, however, is to get you
comfortable with R since we will be using R for geospatial analysis (which is
not so easy in Excel).  The other reason is that with R, you can make your
analysis much more reproducible than you can with Excel or other graphical
software.  In R, everything you do goes into a plain text file called a *script*,
and to rerun your analysis you can simply rerun the same script.  Scripts are
also fairly easy to edit and repurpose (for example, redoing the exact same
analysis with next year's dataset) as well as share with other people (your
neighbor who wants to try your analysis but with his or her own data).

## Dataset for this lesson

As an example of a dataset in spreadsheet form that we can use to demonstrate
the R language, we are using the USDA - Economic Research Service's fertilizer
use and price dataset, available at
https://www.ers.usda.gov/data-products/fertilizer-use-and-price.aspx.
Table 2 of this dataset, showing 1,000 nutrient short tons of nitrogen,
phosphorus, and potash used on corn, soy, cotton, wheat, and other crops in the
US from 1964-2016, has been reformatted into a CSV file for ease of import into
R.  You can download the CSV from (insert link to raw file on GitHub here).  It
can be opened in Excel if you want to get a quick look at it that way, or we can
dive straight into looking at it in R.  Save it into your `data` folder in
your RStudio project for this lesson.

In R, we can use a *function* called `read.csv` to import data from a CSV
(comma-separated value) file. When using any R function, we need opening and
closing parentheses after the function name.  Inside those parentheses we put
items called *arguments* that tell the function what to do.  In this case, we
will put the name of the file to read in quotes, including the folder that it is
in.  (You can use a forward slash regardless of whether you are on Windows or
another operating system.)  To the left of the function name we have the
*assignment operator*, `<-`, which we will use to put the data somewhere for
further processing.  In this case, we will make an *object* called `fert_use`
where we will store the data.

```{r}
fert_use <- read.csv("data/fertilizer_use.csv")
```

Immediately after loading a dataset, it is a good idea to inspect it and make
sure it looks like it is formatted correctly.  The `str` function is helpful
for getting a quick look.

```{r}
str(fert_use)
```

We see that `fert_use` is something called a *data frame*, which is a typical
format in R for storing spreadsheet-like data.  There are five variables,
which we can see are Year, Crop, Nitrogen, Phosphate, and Potash.  Within each
variable there are 265 observations.  We also see that Year, Nitrogen, Phosphate,
and Potash are all represented as integers.  Crop is something called a factor,
which means it is text that is assumed to represent a set of categories.

**Note:**  If you are importing from `read.csv`, `read.table`, or a related
function and you want to treat text columns as text rather than categories,
set the argument `stringsAsFactors = FALSE`.  It is always easy to convert text
to a factor later, and many functions will do this on-the-fly for you.

### Exercise

In either Excel or RStudio's text editor, add a question mark after one of the
numbers in the file.  Re-run `read.csv` and `str`.  What is different?  How
might this be a problem?  Change the file back (or re-download it) when you are
done, then re-run the lines to load in the correct version of the dataset.

**Solution:** When `read.csv` encounters anything in a column that cannot be
formatted as a number, it treats the whole column as text, which by default
causes it to be imported as a factor.  If we want to perform math or make graphs
with numerical data, we won't be able to do so if it is formatted as a factor.

## Functions and vectors

We just used the `read.csv` and `str` functions.  Let's try out some math, now
that we have confirmed that everything that should be a number is a number.
We can use the `mean` function to find the average value of a variable.  How
can we extract one variable from a data frame?  The simplest way is to use
the `$` operator.  It will get us one column of the data frame, which in R
is something we call a *vector*.  `str` works on vectors as well to tell us
something about them.

```{r}
fert_use$Nitrogen
str(fert_use$Nitrogen)
```

We can then pass the whole nitrogen vector to the `mean` function.

```{r}
mean(fert_use$Nitrogen)
```

Whoops!  That is probably not what we wanted.  In R, `NA` means missing data.
You can see from when we printed out the whole vector that we have a lot of `NA`
values.  By default, if `mean` encounters any `NA` values, it returns `NA`.  To
figure out what to do, let's look at the help page.  To find the help page for any
function, type a question mark and then the name of the function.

```{r eval = FALSE}
?mean
```

Ah, there is an argument called `na.rm`, and the description of it says that it
controls whether or not `NA` values are stripped out before calculating the mean.
We can also see that both `na.rm` and `trim` have default values.  We want to
leave `trim` at the default, so we will name `na.rm` when we use it, kind of like
it is named in the Usage section.  In R, `TRUE` and `FALSE` should always be in
all caps.

```{r}
mean(fert_use$Nitrogen, na.rm = TRUE)
```

Note that `fert_use$Nitrogen` got passed to the `x` argument.  If we use
arguments in the exact order they appear in the help file, we don't have to name
them.  But honestly, sometimes naming our arguments just makes code more readable.
And readability is good if you plan on looking at your code again next year
and remembering what you did and why.

```{r}
mean(x = fert_use$Nitrogen, na.rm = TRUE)
```

If we look up `?median`, we can see that there is a similar function for getting
the median of a set of values.  Sometimes the median is a lot more useful than
the mean if there are extreme values in the data set.

```{r}
median(fert_use$Nitrogen, na.rm = TRUE)
```

### Exercise

Is there a function to get the maximum value of a vector?  Try to find the
help page for it, or use Google to find the function name if that fails you.
What are the arguments?  Try finding the maximum amount of phosphate applied
to any crop across all years.

**Solution:** The function is called `max`, and we can view the help page by
typing `?max`.  Like `mean` and `median`, there is an `na.rm` argument to
ignore missing data.  We also see an argument `...`.  Because this is the
first argument, it means you can put in multiple vectors and get the maximum
value across all of them.  We can get the maximum phosphate this way:

```{r}
max(fert_use$Phosphate, na.rm = TRUE)
```

## Arithmetic

A handy feature of R is vectorized arithmetic, meaning we can do arithmetic on
a whole vector at once.  For example, to convert the units for nitrogen application
from 1,000 short tons to short tons, we could multiply by 1000.

```{r}
fert_use$Nitrogen * 1000
```

Works great, and the `NA`s even carried over.

But wait...

```{r}
fert_use$Nitrogen
```

That didn't save our work.  Any time we want to change the value of something in R,
we have to use the `<-` operator.  We probably want to avoid overwriting our
nitrogen column anyway; let's make a new column with a new name.

```{r}
fert_use$Nitrogen_short_tons <- fert_use$Nitrogen * 1000
str(fert_use)
```

Ok, that worked!  The new column is "numeric" instead of "integer", but unless
we're getting really advanced with R, we don't care about the difference between
those.

We can also do arithmetic with multiple vectors.

```{r}
fert_use$All_inputs <- fert_use$Nitrogen + fert_use$Phosphate + fert_use$Potash
head(fert_use)
```

### Challenge

Make a new column called `NP` that is the sum of nitrogen and phosphate applied.
Then make another column called `NP_centered`, which is the value of `NP` minus
the mean of `NP`.  This new column will have negative values if the total amount
of nitrogen and phosphorus applied was less than average, and positive if it was
more than average.

**Solution:**

```{r}
fert_use$NP <- fert_use$Nitrogen + fert_use$Phosphate
fert_use$NP_centered <- fert_use$NP - mean(fert_use$NP, na.rm = TRUE)
head(fert_use)
```

## Creating your own functions

Let's say we want to convert short tons to metric tons so that we can compare
these numbers to some reported from Europe.  Similarly to above, we can do

```{r}
fert_use$Nitrogen_metric_tons <- fert_use$Nitrogen_short_tons * 0.907
```

We can make similar lines for phosphate and potash.  There's nothing
particularly wrong with this, but we can make the code a bit easier to read and
edit by defining a new function for the conversion.

```{r}
ShortTonsToMetricTons <- function(val){
  out <- val * 0.907
  return(out)
}

ShortTonsToMetricTons(5)
fert_use$Nitrogen_metric_tons <- 
  ShortTonsToMetricTons(fert_use$Nitrogen_short_tons)
```

What did I just do there?  I used the `function` keyword to create a new
function, and I used the assignment operator (`<-`) to give that function a
name.  The object `val` is an *argument* for the function, and you'll notice that
neither `val` nor `out` appear in your global environment aftern running the
function.  Between the curly brackets (`{}`), is what we call the
*function body*, which describes what computations the function should perform.
There is also a `return` statement that shows what the function *returns*,
*i.e.* what the output should be.

Now, when reading back through your code, you don't have to remember what the
number 0.907 was for; you have a function name that describes what it does.
Additionally, say you later decide that you want to make the conversion with
more precision.

```{r}
ShortTonsToMetricTons <- function(val){
  out <- val * 0.907185
  return(out)
}
```

There's only one place that you have to edit the number, even if you use the
function dozens of times in your code.  That's a lot less error-prone than
going back and editing the number in dozens of places.  You still have to
re-run the code to change all of the results, but at least the editing is
minimal.

That was a relatively simple example, but later when you write much larger
chunks of code and want to avoid having to copy and paste it all, functions
can be very helpful.  We'll make a few functions in later episodes of this
lesson, which is why we at least wanted to introduce the concept here.

## Installing packages

There are over 15,000 add-on packages for R, all completely free.
CRAN is the official place where they are hosted, and you can see a list of
CRAN packages here:
https://cloud.r-project.org/web/packages/available_packages_by_name.html

Additionally, many people who create small R packages and don't want to deal
with the hassle of hosting them on CRAN will make their package available
via GitHub.  If you really get into R and want to share your code with others,
you might even make your own R package one day!

In the meantime, let's see how to get R packages from CRAN.  For this episode,
we will use a package called `ggplot2`.  The function for installing packages
is called `install.packages()`.

```{r eval = FALSE}
install.packages("ggplot2")
```

Once a package has been installed, we don't need to install it again until we
upgrade R.  To avoid accidentally reinstalling, this is a good time to
"comment out" our code.

```{r}
#install.packages("ggplot2")
```

If we want to use a particular package in our script, we also need to load it
with `library()`.  We won't comment that out, since we will need to do it
every time we reopen R and run the script.  Generally, it is good to put
all calls to `library()` at the top of the script, so that as soon as you
(or someone else) open the script, you know what you need to have installed.

```{r}
library(ggplot2)
```

(It's one of R's little quirks that package names need to be in quotes for
installation, but not for loading.)

### Exercise

Later today we will use the package `rgdal`.  Install it and load it.

**Solution:**

```{r eval = FALSE}
install.packages("rgdal")
library(rgdal)
```

## Graphics with ggplot2

Although there are other packages for making graphics, `ggplot2` is popular
for being easy to learn and versatile.  We'll scratch the surface of it in this
lesson.  To learn more, go to https://ggplot2.tidyverse.org/.

Let's plot nitrogen applied vs. year.  To start, we will pass our data frame
to the `ggplot` function as the first argument.  The second argument is a set
of *aesthetics* indicating how variables should correspond to various aspects
of the plot.

```{r}
ggplot(fert_use, aes(x = Year, y = Nitrogen))
```

This sets up the plot but doesn't put anything in it.  We can use `geom_point` to
make a scatter plot.  We will use the `+` operator to connect these commands.

```{r}
ggplot(fert_use, aes(x = Year, y = Nitrogen)) +
  geom_point()
```

Looks like some crops changed over time and others didn't.  But which ones?

```{r}
ggplot(fert_use, aes(x = Year, y = Nitrogen, color = Crop)) +
  geom_point()
```

Maybe we want to add some lines connecting the years.  In the help page for
`geom_line`, we see the `group` aesthetic, which we will also need here.

```{r}
ggplot(fert_use, aes(x = Year, y = Nitrogen, color = Crop, group = Crop)) +
  geom_point() +
  geom_line()
```

We can also make separate plots for each crop using the `facet_wrap` function.
Since some crops used a lot less nitrogen than others, we'll set
`scales = "free_y"` to get a better picture of how each crop changed over time.
Notice also that we need the `~` symbol with `facet_wrap`.

```{r}
ggplot(fert_use, aes(x = Year, y = Nitrogen, color = Crop, group = Crop)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ Crop, scales = "free_y")
```

### Exercise

Make a scatter plot of nitrogen vs. potash, and color it by crop.  Make
another plot where instead of coloring by crop, you facet by crop.
Now try it with `facet_wrap(~ Year > 1980)`.  What happens?

**Solution:**
```{r warning = FALSE}
ggplot(fert_use, aes(x = Potash, y = Nitrogen, color = Crop)) +
  geom_point()

ggplot(fert_use, aes(x = Potash, y = Nitrogen)) +
  geom_point() +
  facet_wrap(~ Crop)

ggplot(fert_use, aes(x = Potash, y = Nitrogen, color = Crop)) +
  geom_point() +
  facet_wrap(~ Year > 1980)
```

