---
title: "Ag Carpentry - Weather and Soil Data"
author: "Brittani"
date: '2019-10-29'
output:
  html_document: default
  pdf_document: default
source: Rmd
---

```{r setup, include=FALSE}
library(knitr)
library(sf)
library(fasterize)
library(raster)
library(rjson)
library(httr)
library(rgdal)
library(rgeos)
library(maptools)
library(knitr)
require(tmap)
require(ggplot2)
require(gridExtra)
library(daymetr)
library(readr)
library(measurements)
library(FedData)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE, fig.path='../fig/')
source("functions.R")
```
####Motivating Questions:
- What are the common file types in agricultural data?
- What publicly available datasets exist for my field?

####Objectives with Agricultural Data Types
- Describe the format of public weather and soil datasets
- Import weather data from the internet, eg. daymetr
- Access to elevation and ssurgo data with higher resolution
- Derive topography data from elevation data

####Keypoints:
- sf is prefereable for data analysis; it is easier to access the dataframe
- Projecting your data in utm is necessary for many of the geometric operations you perform
(e.g. making trial grids and splitting plots into subplot data)
- Compare different data formats, such as gpkg, shp(cpg,dbf,prj,sbn,sbx),geojson,tif

> # Daymet Weather Data
> 
> The Oak Ridge National Laboratory produces a datset called Daymet which contains predicted
> weather observations on a one meter grid. These data come from weather station climate 
> observations in a climate model for prediction and include variables such as precipitation,
> snow water equivalent, temperature, day length, solar radians, and vapor pressure. 
> 
>There is a package in r `daymetr` that downloads the daymet weather data within the R environment.
> For a single point, you can use the command `download_daymet()`. If you want to download the data 
> for a set of points, there is also the command `download_daymet_batch()` which takes an argument 
> a .csv of the points in lat/long.
> 
{: .callout}

> # Using daymetr to download weather
> We will use the mean latitude and longitude values from the bounding box as our point for the 
> weather data. The functions `cent_long()` and `cent_lat()` give the mean longitude and latitude points of the bounding box around the `sf` object that is supplied. This should be a point near the middle of the field. 
> 
> ```{r location}
> boundary <- read_sf("data/boundary.gpkg")
> lon <- cent_long(boundary)
> lat <- cent_lat(boundary)
> ```
>
> We also call the site `Field1`, but this can be the name of one of your fields if you use it in the future.
> We can choose the start and end years. If the data is not available for the year you request, an 
> error will be reported. We choose 2000 to 2018 for this example; later we will use the
> historical data for comparison. The final option `internal = TRUE` means that the daymet 
> data is brought into the R environment rather than saved in your working directory.
>
>```{r daymetr}
> weather <- download_daymet(site = "Field1", lat = lat, lon = lon, start = 2000, end = 2018, internal = TRUE)
>```
{: .callout}

> # Exploring the daymet data
> We can use `str()` as we did in the last lesson to explore the new datmetr object `weather`
> ```{r}
> str(weather)
>```
> The object `weather` is a list of 7 objects, the last of which is the data. In the next excerise we will explore what variables are in this dataframe. 

> # Exercise 1: Explore the weather data
>
>  1. Grab this object and save it as `weather_data`.
>  2. How is the date reported? 
>  3. What other variables exist?
>  4. What are the units for the different variables?
>  *Remember: Sometimes you need to use a search engine to understand what objects
>  are created from a specific R function. 
>
> > #Solution
> > 
> >```{r weatherdata}
> >weather_data <- weather$data
> >str(weather_data)
> >```
> > The date is reported as the year and day of the year. 
> > Other variables include day length,
precipitation, solar radiation, snow water equivalent, maximum temperature, minimum
temperature, and vapor pressure. 
> > The units for the variables are given after the variable 
name. For example, day length is in seconds and solar radiation is in watts per square meter.
While precipitation and temperature have intuitive names, vapor pressure and snow water
equivalent are not so apparent. 
> > Use the `daymetr` [vignette](https://cran.r-project.org/web/packages/daymetr/vignettes/daymetr-vignette.html) to understand the meaning of these variables.
> > 
> {: .solution}
{: .challenge}

> #Dates in Dataframes 
>
> There is a class within R for dates. Once a column is of the `date` class, we can subset
> or order the dataset by time. `as.Date()` converts a column to a data, but here if we try 
> the command `weather_data$date <- as.Date(weather_data$yday)`, we will receive an error 
> saying an origin must be supplied. 
>
> The function can see that the date is in days after some starting time or origin. The name
> `yday` means this is the day of the year, so the origin should be the last day of the 
> previous year. There are multiple years in our dataframe, so the origin should change 
> for each year. This is accomplished in the function as.Date.daymetr(). To see this code and understand how > it works go to the [functions script](https://github.com/data-carpentry-for-agriculture/trial-lesson/blob/gh-pages/_episodes_rmd/functions.R) in github. 
> 
> ```{r dateformat}
> weather_data$date <- as.Date.daymetr(weather_data)
> head(weather_data$date)
> class(weather_data$date)
> ```
{: .callout}

> #Unit Conversions
>
>Publicly available data are usually given in metric units as we saw in the weather data 
>above. We may want to have these data in imperial units as these are the units we are using
>to think about yield and other values in the United States. You may know the value of crop requirements and threshholds in imperial units rather than metric units. For example, a goal of 18 inches of rain during the corn season.
>
>The package `measurements` is used for converting different types of measurements from one unit to another. The 
> command `conv_unit()` converts the column from one stated unit to another unit.
> Another useful function in the package is `conv_unit_options` which gives the possible units for a specific kind of measure (e.g. length, area, weight, etc.). 
> 
> We have made simple functions for converting units within `conv_unit()`. For example, the function `mm_to_in()` can convert the daily precipitation from milimeters to inches. The following lines converts `prcp..mm.day.` to inches and creates a new column called `prec`.
> 
> ```{r convprec}
> weather_data$prec <- mm_to_in(weather_data$prcp..mm.day.)
> 
```

> Exercise 2: Unit Conversions
>
> 1. Convert the two temperature variables into fahrenheit from celsius using the function `c_to_f()` with the names `tmax` and `tmin`. 
> 2. What is the maximum temperature recorded?
>
>> Exercise 2 Solutions
>> 
>> ```{r exercise1}
>> weather_data$tmax <- c_to_f(weather_data$tmax..deg.c.)
>> weather_data$tmin <- c_to_f(weather_data$tmin..deg.c.)
>> head(weather_data$tmax)
>> max(weather_data$tmax)
>> ```
>> The maximum temperature during this time period was 94 degrees.
> {: .solution}
{: .challenge}


> #Precipitation Graph
>
> Perhaps you want to see what the weather this year was like compared to the typical year.
> We are going make a graph showing the total monthly precipitation from
> 2018 compared to the average precipitation from the years 2000 to 2017. This is a common way
> to look at seasonal rainfall and allows us to look at the rainfall during the critical months
> of July and August. First, we need to organize the data in a new dataset with the average total rainfall for each month.

> # Adding a month variable
> 
> Currently, there is no month variable in our dataframe. There is a package called `lubridate`
> that can facilitate easy transformations of dates in R. We use the command `month()` to add a 
> variable called `month` to the dataframe. The option `label = TRUE` creates a string with the 
> month name instead of a number. 
> 
> ```{r addmonth}
> weather_data$month <- lubridate::month(weather_data$date, label = TRUE)
> head(weather_data$month)
> ```

> # Taking sum or average by groups 
>
> Now, we need to sum the daily precipitation for each year and month combination. There is a 
> package called `dplyr` that helps with many kinds of data manipulation. A popular task is to 
> perform an action over a group, like taking the sum of something. To specify the grouping variables, you use `group_by()` then add the additional
> command `summarise()` which defines the action. For this exercise we wrote functions that use `dplyr` to make the task simpler for you 
> `sumprec_by_monthyear()` and `avgprec_by_month()`.
> 
> First we use the command to calculate the total precipitation for each month in each year.
> ```{r monthyeardata}
> by_month_year <- sumprec_by_monthyear(weather_data)
> head(by_month_year)
> ```
>
>Now we have a dataframe with the rainfall for each month of each year, where the first rows of the dataframe are for January.
>


Now we want to separate the 2018 data from the rest of the years before we take the average monthly precipitation. We will take the subset of `by_month_year` that is in 2018.

```{r monthdata}
monthprec_2018 <- subset(by_month_year, year==2018) 
```

Now find the subset of `by_month_year` that is not in 2018, and we will take the average of the precipitation for each month. 

```{r monthdatahist}
monthprec_hist <- avgprec_by_month(subset(by_month_year, year != 2018))
```

We now have two separate dataframes `by_month_2018` and `by_month` with the rainfall for each
month. We can use the common variable `month` to merge them into one dataframe with the average
monthly rainfall and the 2018 monthly rainfall using the `merge()` function.

```{r plotdata}
prec_plot <- merge(monthprec_hist, monthprec_2018, by = "month")
```

We will now use `ggplot` to create a graph with a bar representing the monthly precipitation 
in 2018 and a point with the average rainfall from 2000 to 2017. In the function 
`geom_bar()` `stat = identity` creates a bar graph where the height of the bar is the value of the 
variable rather than the count of the observations, the common use of a bar chart. 

```{r precmonthplot}
monthly_prec <- ggplot(prec_plot) + 
  geom_bar(aes(x = month, y = prec_month), stat = 'identity') 
monthly_prec + geom_point(aes(month, prec_avg), show.legend = TRUE) + ggtitle("Field 1")
```

The most notable feature of the weather graph is the below average rainfall in July, the most 
critical growing period for corn. To understand whether this affected yield on the field, we
woud also need to look at historic yield. But on your field, you will know those historic
average and be able to have a pretty clear idea of weather impacted the average yield in
a growing season. 

Another possible graph you could make with these data is on the accumulated GDD each month. 

