---
title: "Ag Carpentry - Experimental Design"
author: "Aolin Gong"
include_overview: true
output: html_document
questions:
 - What kind of on-farm experiments do we do?
 - How do we design these experiments efficiently?
objectives:
 - Know different types of common on-farm experiments
 - Import boundary file and AB line file
 - Create AB line file with code
 - utilize the functions to create simple trial designs
keypoints:
 - Most of the code in this part would be using the functions, therefore understanding what different functions can be quite important
 - In designing the trials, the most important thing is to know how to design the experimental rates,and the tech part can be done by someone else
source: Rmd
---

```{r setup, include=FALSE}
library(knitr)
library(sf)
library(fasterize)
library(raster)
library(rjson)
library(httr)
library(rgdal)
library(rgeos)
library(maptools)
library(knitr)
require(tmap)
require(ggplot2)
require(gridExtra)
library(daymetr)
library(readr)
library(measurements)
library(FedData)
library(lubridate)
library(data.table)
library(dplyr)
library(tidyverse)
library(tidyr)
library(broom)
source("../bin/chunk-options.R")
# source("function.R")
```

<font color="magenta">Words about what we are actually doing go here</font>

The only files we need for the trial design are the boundary file, and ab line. In addition, as long as we know the actual direction the machines
will be driven on the field, we can create our own AB line with a function.

```{r read_files}
boundary.sfdf <- st_read("data/boundary.gpkg")

boundary.utm <- st_transform_utm(boundary.sfdf)

```



#####Field Management Info Above
We need decide on the experiment design before we get into any of the code, relative parameters we need for the trial design includes: plot dimension, number of treatments, types of treatments, and treatment range. In the following code, we are simply going to assign values to all
the parameters that might be involved in the trial design. In this way, if we ever want to change any parameters, we can do it here, and need not
to worry about the consistency for the whole code.
```{r parameter_setting}
num_treats <- 5  #number of treatments
#--- direction ---#
# pick one of c('SN','NS','WE','EW')
# the direction in which the applicator moves first
# SN means the applicator goes from south to north 
long_in <- 'NS' 
short_in <- 'EW'  
#--- grid parameters in feet ---#
plot_length_ft <- 30
plot_width_ft <- 30 #usually this is determined by width of the equipments
#--- meter to feet ---#
feetinameter <- 3.28084
#--- grid parameters in meter ---#
plot_length_meter <- plot_length_ft/feetinameter
plot_width_meter <- plot_width_ft/feetinameter

LongA <- -82.87339
LatA <- 40.84575

LongB <- -82.87329
LatB <- 40.83987


bothfields <- subset(boundary.utm, Type == "Trial")
bbox_field <- st_bbox(bothfields)
# +c(100,100,50,50)
xmin <- bbox_field[1]
ymin <- bbox_field[2]
xmax <- bbox_field[3]
ymax <- bbox_field[4]

#--- identify the starting point ---#
starting_point <- c(xmax,ymax) 
 
  
seed_treat_rates <- c(24000, 27000, 31000, 34000, 37000, 40000)
nitrogen_treat_rates <- c(130,160,200,225,250,275) 
seed_quo <- seed_treat_rates[3]
nitrogen_quo <- nitrogen_treat_rates[3]
num_treats <- 25
```


###make our own AB line if you do not already have one
```{r create an AB line}
ab_line <- make_abline(LongA, LongB, LatA, LatB, as.numeric(paste0("326", "17"))
)
```

eg: read your own ab line file

eg create plot to see the boundary and abline
tm_shape(bothfields) + tm_borders(col='green') + tm_compass() + tm_grid() +
  tm_shape(ab_line) + tm_lines(col='red', scale=3)


st_crs(bothfields)
st_crs(ab_line)
#st_crs(starting_point) <- st_crs(bothfields)

```{r design the checker-board plot}
design_grids_utm <- make_grids(bothfields, abline_utm, long_in, short_in, length_ft = 30, width_ft = 30)

st_crs(design_grids_utm) <- st_crs(bothfields)

trial_grid1_utm <- st_intersection(bothfields, design_grids_utm)

tm_shape(trial_grid1_utm) + tm_borders(col='blue')

```

eg 
tm_shape(bothfields) + tm_borders(col='blue') + 
  tm_shape(design_grids_utm) + tm_borders(col='pink')+ tm_text('plotid', size = 0.5)

st_crs(design_grids_utm)<-st_crs(bothfields)

tm_shape(trial_grid1_utm) + tm_borders(col='blue') + tm_text('plotid', size = 0.5)

```{r clean up plots}
max_area <- as.numeric(mean(st_area(trial_grid1_utm)))

trial_grid2 <- trial_grid1_utm %>%
  mutate(area=as.numeric(st_area(.))) %>%
  mutate(drop=ifelse(area<(max_area), 1, 0) )

trial_grid2_intrial <- subset(trial_grid2,drop==0) #the plots that are in the trial--not too small,
#and not in the buffer zone.
```


eg make plots to show the grid
tm_shape(bothfields) + tm_borders(col='blue') + 
  tm_shape(ab_line) + tm_lines(col='red', scale = 4) +
  tm_shape(trial_grid2) + tm_polygons('drop') + tm_text('plotid', size = 0.5)


### determine treatment types
Now that we have the trial design plots, we need to assign different treatments to each plot.
```{r assigning treatment rates}
whole_plot <- treat_assign(trial_grid2_intrial,num_treats,seed_treat_rates,nitrogen_treat_rates,seed_quo,nitrogen_quo)
```
plot(newfield$geom)

```{r cleanup}
infield <- st_buffer(boundary.utm,-plot_width_meter)
outfield <- st_difference(boundary.utm,infield)

infield$dummy <- 1
outfield$dummy <- 0
bothfields <- rbind(infield, outfield)
newfield <- subset(bothfields, dummy == 1)



```


###eg plot
  
ggplot(data = whole_plot, aes(x=long,y=lat,group=group)) +
  geom_polygon(aes(fill=factor(NRATE))) +
  scale_fill_brewer(palette = 'Greens') 

