---
title: "Experimental Design"
include_overview: true
output: html_document
questions:
 - What kind of on-farm experiments do we do?
 - How do we design these experiments efficiently?
objectives:
 - Know different types of common on-farm experiments
 - Import boundary file and AB line file
 - Create AB line file with code
 - utilize the functions to create simple trial designs
keypoints:
 - Most of the code in this part would be using the functions, therefore understanding what different functions can be quite important
 - In designing the trials, the most important thing is to know how to design the experimental rates,and the tech part can be done by someone else
source: Rmd
---


```{r setup, include=FALSE}
library(knitr)
library(sf)
library(fasterize)
library(raster)
library(rjson)
library(httr)
library(rgdal)
library(rgeos)
library(maptools)
library(knitr)
require(tmap)
require(ggplot2)
require(gridExtra)
library(daymetr)
library(readr)
library(measurements)
library(FedData)
library(lubridate)
library(data.table)
library(dplyr)
library(tidyverse)
library(tidyr)
library(broom)
source("../bin/chunk-options.R")
source("functions.R")
```

> ## Trial Design
> Now we will design our own experiments on the field. The only files we will need for the trial design are the boundary file and ab line.
> 
{: .callout}

> ## Read and transform shape files
> 
> We will start by reading in the shape files we need like we've been doing for the last few episodes:
> ```{r}
> boundary <- st_read("data/boundary.gpkg") # read in boundary
> abline <- st_read("data/abline.gpkg") # read in AB line
> ```
> 
{: .callout}

> Now let's check the coordinate references of our two files:
> ```{r}
> st_crs(boundary)
> st_crs(abline)
> ```
> Since both of these are in lat/long and we want them in UTM, we'll transform them:
> 
> ```{r}
> trialarea <- st_transform_utm(boundary)
> abline_utm <- st_transform_utm(abline)
> ```
> 
{: .callout}

> ## Designing trials
> 
> We need decide on the details of experiment design before we get into any of the code.  Relative parameters we need for the trial design include:
>  * plot dimensions
>  * number of treatments
>  * types of treatments, and
>  * treatment range.
> 
{: .callout}

> ## Defining Parameters
> In the following code, we are simply going to assign values to all the parameters that might be involved in the trial design. In this way, if we ever want to change any parameters, we can do it here, and need not to worry about the consistency for the whole code.
> 
> Now let's design our grid with the following parameters:
> ```{r}
> width_in_meters = 24 # width of grids is 24 meters
> long_direction = 'NS' # direction of grid that will be long
> short_direction = 'EW' # direction of grid that will be short
> length_in_ft = 180 # length of grids in feet
> ```
> 
{: .callout}

> ## Make Grids
> We'll use our `make_grids` again function to generate this trial's grid:
> ```{r}
> width <- m_to_ft(24) # convert meters to feet
design_grids_utm <- make_grids(trialarea, abline_utm,
             long_in = long_direction,
			       short_in = short_direction,
			       length_ft = length_in_ft,
			       width_ft = width)
> ```
> Next we want to make sure the coordinate reference frame of our `trialarea` is the same as our `design_grids_utm` grids and then take the intersection of these grids with our trial area as we did previously:
> ```{r}
> st_crs(design_grids_utm) <- st_crs(trialarea)
> trial_grid <- st_intersection(trialarea, design_grids_utm)
> ```
> 
> Let's check out what our trial subplots look like:
> ```{r t3jpn}
> tm_shape(trial_grid) + tm_borders(col='blue')
> ```
> 
{: .callout}

> ## Determining subplot treatments
> 
> Now that we have the trial design plots, we need to assign different treatments to each plot.  We can use the `treat_assign` function from `functions.R` to randomly assign seed rates and nitrogen rates to each plot on our grid.
> 
> We'll select 4 different seed rates and 4 different nitrogen rates to deposit randomly on our grid:
> ```{r}
> seed_rates <- c(31000, 34000, 37000, 40000)
> nitrogen_rates <- c(160,200,225,250)
> ```
> The `seed_quo` and `nitrogen_quo` are the rates that will be applied to the headlands that are not part of the trial.
> ```{r}
> seed_quo <- 37000
> nitrogen_quo <- 225
> ```
> 
{: .callout}

> # Generating Treatment Map
> We are now ready to generate our treatment plot:
> ```{r message=FALSE, warning=FALSE}
> whole_plot <- treat_assign(trialarea, trial_grid, head_buffer_ft = width,
                           seed_treat_rates = seed_rates,
			   nitrogen_treat_rates = nitrogen_rates,
			   seed_quo = seed_quo,
			   nitrogen_quo = nitrogen_quo)
```
> 
{: .callout}

> ## Mapping Trial
> Let's look at what our trial looks like.  First, as a shape file:
> ```{r}
> head(whole_plot)
> ```
> And as plots:
> ```{r t2jpn}
> nitrogen_plot <- map_poly(whole_plot, "NRATE", "Nitrogen Treatment")
> seed_plot <- map_poly(whole_plot, "SEEDRATE", "Seedrate Treatment")
> treatment_plot_comp <- tmap_arrange(nitrogen_plot, seed_plot, ncol = 2, nrow = 1)
> treatment_plot_comp
> ```
> 
{: .callout}

<!-- Test
Brittani testing whether I'm the problem with the rendering
-->
<!-- JPN takes out
```{r assigning treatment rates, message=FALSE, warning=FALSE}
whole_plot <- treat_assign(trialarea, trial_grid, head_buffer_ft = width, seed_treat_rates = c(31000, 34000, 37000, 40000), nitrogen_treat_rates = c(160,200,225,250), seed_quo = 37000, nitrogen_quo = 225)
map_poly(whole_plot, "NRATE", "Nitrogen Treatment")
```
```{r}
head(whole_plot)
```
The function will automatically take out 

```{r cleanup}
# infield <- st_buffer(boundary.utm, -plot_width_meter)
# outfield <- st_difference(boundary.utm, infield)
# 
# infield$dummy <- 1
# outfield$dummy <- 0
# bothfields <- rbind(infield, outfield)
# newfield <- subset(bothfields, dummy == 1)
```


###eg plot

```{r}
ggplot(data = whole_plot, aes(x=long,y=lat,group=group)) +
  geom_polygon(aes(fill=factor(NRATE))) +
  scale_fill_brewer(palette = 'Greens') 
```
-->
